---
title: "Cancer methylome analysis CLL vs. B-Cells"
author: "Leona Brandl, Carlotta Brüggen, Tim Kühn, Violetta Schaaf"
date: "8th July 2019"
output: html_document
---


Text about what our project is about.

The analysis will be done in several steps:


1. **Data processing**
  * Reorganisation of the data
  * Quality control 
  concerning NAs, coverage values and unusefull DNA-regions
2. **Normalisation**
  * Transformation of beta-values into M-values
  * Quality control of M-values
3. **Data reduction**
  * Dimensionality reduction via PCA
  * Featuere selection
4. **Clustering**
  * Via K-means clustering
   
5. **Identifying DMRs**
6. **logistic regression**
7. **Data interpretation**


### Data processing

Before we can start analysing our data set it has to be tidied up to make it easy for us to work with it by making it as clear as possible. Datasets have to be split up and reorganized, coloums removed or renamed and a quality control concering coverage values, beta values, NAs and specific DNA segments showing unusual values, needs to be performed.

#### Reading in the datasets

We start by **reading in our CLL-Bcell-data and the annotations** which is an extra document explaining where our data is coming from.

```{r}
input_data <- readRDS(file ="CLL-Bcells_list.RDS.gz")

annotation <- read.csv("sample_annotation.csv")
```

#### Reorganisation of the data

Our main dataset, the one containing the data we are working with, is `input_data`. It contains methylation data in form of beta values and coverage values of B-cells of five healthy patients and five patients suffering from CLL. It is divided into four subgroups: tiling, genes, promoters and CpG islands. We have to analyse those subgroups separatly, because methylation in promoters, genes and CpG islands can have different biological reasons and also this way the data the computer deals with is not that much while running the algorithm. Therefor **the dataset gets divided into four subdatasets**.

```{r}
genes <- input_data$genes

promoters <- input_data$promoters

cpgislands <- input_data$cpgislands

tiling <- input_data$tiling
```

Those subgroups datasets contain methylation data for different regions on the genome:
`genes`, `promoters` and `cpgislands`contain genome segments with specific functions and methylation features. Genes, especially highly expressed genes, sometimes show methylation in their gene body. Unfortunately the biological function behind this is not fully clear yet. Still it can be interesting investigating in their methylation pattern in terms of cancer. In promoters  we already know that methylation plays an important role, because as promoters are the starting point for transcription, their methylation can silence genes laying downstream of these promoters. This way for example tumor suppresor genes don't get expressed any more, which can lead to cancer. Here we expect to find crucial differences in methylation between sick and healthy cells. CpG islands also are important regions on the DNA with a high density of cytosines in neighbourhood to guanines. These CpG islands are often in promoter regions of housekeeping genes and in healthy beeings only rarely methylated, because their methylation usually leads to mistakes in replication and transcription and quickly in death just because functioning promoters are so essential in those genes. Also in those CpG islands a loss of function can lead to cancer, so we expect significant methylation differences. The fourth subgroup is `tiling` which describes a specific section on the genome in steps of 5000 nucleotides without a biological context.

As we want to analyse those subgroups separatly, we first just choose one subgroup, `genes`, 
to work on our code. Later we can transfer the code on the other subgroups, too.
In this `genes`dataset we **rename the columns**, which stand for different patients (patient 1 to patient 10). They get short precise names to let the reader get the meaning on the first sight.

```{r}
colnames(genes)[11:30] <- c("P1_healthy_beta","P2_healthy_beta","P3_healthy_beta","P4_healthy_beta","P5_healthy_beta","P6_CLL_beta","P7_CLL_beta","P8_CLL_beta","P9_CLL_beta","P10_CLL_beta","P1_healthy_coverage","P2_healthy_coverage","P3_healthy_coverage","P4_healthy_coverage","P5_healthy_coverage","P6_CLL_coverage","P7_CLL_coverage","P8_CLL_coverage","P9_CLL_coverage","P10_CLL_coverage")
```

####quality control

It is important to remove values that are not needed for or even hinder further analysis. Among other values those are also coverage values, which seem to be unrealistically high due to PCR dublicates or repetetive regions. Also they can be or that low, so the belonging beta-value of the DNA section has no meaning, because there was no fitting allignment when mapping back PCR fragments onto the genome. 
A **data frame only containing coverage values** is created.

```{r}
cov_genes <- genes[,c(1,21:30)]
```

At this point we already delete some genes, which are the genes on the X- and Y-chromosome. Somehow there must have been problems in obtaining the beta and coverage values for those chromosomes, because the beta-values are either not a number (NaN) or mainly 1.000, which must definitely be a mistake.The belonging coverage is in most cases 0, so we can not use those beta-values values anyway. Following code **removes the X- and Y-chromosome**. 

#####Removing genes on X- and Y-chsomosome

```{r}
cov_genes_new <- cov_genes[-which(cov_genes =="chrX"),]
cov_genes <- cov_genes_new[-which(cov_genes_new =="chrY"),]
rm(cov_genes_new) 
cov_genes <- cov_genes[ ,c(2:11)]
```

#####Finding a coverage value threshold

But back to the very high or low coverage values: There need to be a threshold defining a senseful coverage value range. To determine a threshold we have a look at the coverage value distribution among all patients using a logarithmic density plot of the coverage means of all patients for each genome segment. Therefor a **matrix containing the means** in generated.
```{r}
cov_genes_means <- rowMeans(data.matrix(cov_genes))
```
A **logarithmic density plot** using this matrix values is generated.
To **find an upper threshold** for deleting coverage values we try quantile values we read about in literature and draw the lines for those quantiles in the logarithmic plot to see wether one of those coincidences with a kink in the curve. For the **lower threshold** we draw lines at the values 10 and 15, which are common threshold values as we read in litearture.

```{r}
plot(density(log10(cov_genes_means)), xlab = "coverage means", main = "coverage distribution")
quantile(cov_genes_means, probs = c(.95))
abline(v=log10(52529.75))

quantile(cov_genes_means, probs = c(.975))
abline(v=log10(82556.55))

quantile(cov_genes_means, probs = c(.999))
abline(v=log10(311230.4))

abline(v=log10(10))
abline(v=log10(15))
```

Both lower thresholds and the highest upper threshold seem to be fitting looking at the diagram. But which threshold we will chose eventually also depends on the percentage of deleted rows at the end of quality control. It shouldn't be deleted more than 15 % of the rows, because PCA will reduce the amount even more and we don't want to lose too much information.

The procedure will now be to set all the coverage values in `cov_genes`to NA, that are below or above the chosen threshold. Then all the beta-values belonging to those coverage values set to NA will be set to an NA, too. Eventually we will set another threshold for the beta-values, defining above which number of NA per row (meaning gene), the row will be deleted. Therefor we will need two nested for-loops.

#####Preparing the first nested for-loop

Before we can set a threshold, the few **NAs among the coverage values need to be set to 0**. This way they will be set to NAs again after the loop, but they also don't hinder the loop from working.

```{r}
sum(is.na(cov_genes))

cov_genes[is.na(cov_genes)] <- 0
```

A **threshold can be chosen**. As the lower threshold we choose the coverage value 15, which is positioned at a kink of the coverage distribution diagram. As the upper coverage threshold we choose the 90 % quantile. Although it is not positioned at a kink in the diagram we consider it to be a good threshold, because it still only leads to a removal of 7(?) % in the end of the quality control, which is far underneath the upper limit of 15 %.

```{r}
threshold1 <- 15

threshold2 <- quantile(cov_genes_means, probs = seq(0.90,0.90,0.05))
```

#####The first nested for-loop

Now a **nested for-loop** can set the coverage values underneath and above the set threshold to NA.
This will be done for each row and each column of the dataframe `cov_genes`.

```{r}
for(i in 1:nrow(cov_genes)){
  for(j in 1:ncol(cov_genes)){
    if(isTRUE(cov_genes[i,j] <= threshold1)){
      cov_genes[i,j] <- NA
    } 
    if(isTRUE(cov_genes[i,j] >= threshold2)){
      cov_genes[i,j] <- NA
    }
  }
}

rm(i,j,threshold1,threshold2)
```

#####Preparing the second nested for-loop

A second nested for-loop is used to set the beta-values to NA that belong to the coverage values set to NA. Also here a **dataframe containing only beta-values** is created and the **genes located on the X- or Y-chromosome get removed**.

```{r}
beta_genes <- genes[,c(1,11:20)]

beta_genes_new <- beta_genes[-which(beta_genes =="chrX"),]
beta_genes <- beta_genes_new[-which(beta_genes_new =="chrY"),]
rm(beta_genes_new)
beta_genes <- beta_genes [ ,c(2:11)]
```

#####The second nested for-loop

This **second nested for-loop** checks every row and every column of the dataframe `cov_genes`.
If there is an NA, it sets the belonging beta-value in `beta_genes` to NA, too.

```{r}
for(k in 1:ncol(beta_genes)){
  for(l in 1:nrow(beta_genes)){
    if(isTRUE(is.na(cov_genes[k,l] == TRUE))){
      beta_genes[k,l] <- NA
    } 
  }
}
rm(k,l)
```

#####Finding a threshold for NAs among the beta-values

Now the second threshold in quality control needs to be chosen. This one is for the upper and lower threshold of allowed NAs per row (gene) among the beta-values.
Since some beta-values were set to NA in the last step, we want to get insight in **how many NAs we actually have now**.

```{r}
rmv.rows_beta_genes = apply(beta_genes,1, function(x){sum(is.na(x))})
```

A new dataframe containing only beta-values for genes which are inside the threshold is created.
We chose a threshold of 3 allowed NAs per row (gene), so **every row containing more than 3 NAs gets removed**.

```{r}
beta_genes_cleaned <- beta_genes[-which(rmv.rows_beta_genes >2),]
```

To see **how many rows are gone** due to quality control the numer of rows before and after get compared. Also we hava a look at the **percantage of rows deleted** and see wether it is beneath 15 %.

```{r}
row_difference = nrow(genes)-nrow(beta_genes_cleaned)
sum(row_difference)
genes_deleted_percentage = row_difference/nrow(genes)*100
sum(genes_deleted_percentage)
```

With 7% we are beneath the recomended upper limit of 15 % of deleted rows.

###Normalisation

The tidy beta-values we have now are an approximation of the percentage of DNA-methylation. Biologically beta-values are easy to understand. They range from 0 to 1, while 0 is unmethylated and 1 is fully methylated. But their bounded range is also a disadvantage. It leads to the problem that they can not be used for statistical tests like the t-test, because they violate the Gaussian distribution. For highly methylated gene regions and unmethylated gene regions beta-values are very heteroscedastic, which means, that the variability of a variable is unequal across the range of values. Therefor we do a normalisation. Beta-values need to be transformed into M-values via a logit transformation. M-values do not have a bounded range, they can be infinite big and small. While the middle methylation range (beta-value range of 0,2-0,8) is linear to M-values, outside this range the beta-values are compressed and the M-values more accurate and homoscedastic. This is why M-values can be used for statistical tests and are absolutely necessary for further analysis.

#####Preparing beta-values for logit transformation

Logit transformation turns the extreme beta-values 0 and 1 into -infinite and infinite. We can not work with those values, because further tests don't know how to deal with an infinite number. Due to that we apply a little trick and **set 0 to a very small number higher than 0, and 1 to a very big number less than 1**.

```{r}
beta_genes_cleaned[beta_genes_cleaned==0]<-0.00000001
beta_genes_cleaned[beta_genes_cleaned==1]<-0.99999999
```


Because we only want to work with our beta-values we **remove the first column of beta_genes_cleaned which containes the chromosoms.**

```{r}
beta_genes_cleaned <- beta_genes_cleaned[,c(2:11)]
```

#####Creating two seperate data frames for sick and healthy patients

For doing our Normalisation were we transform our beta-values into M-values we have to **split up our data into beta-values of healthy ans sick patients.**

```{r}
beta_genes_healthy <- beta_genes_cleaned[,c(1:5)]
beta_genes_cancer <- beta_genes_cleaned[,c(6:10)]
```

Now we are able to **replace the remained NA's by the row means of the different genes.** We suspect that the rowmeans differ between healthy and sick patients so it is important to work with the two dataframes.


```{r}
k <- which(is.na(beta_genes_healthy), arr.ind=TRUE)
beta_genes_healthy[k] <- rowMeans(beta_genes_healthy, na.rm=TRUE)[k[,1]]

l <- which(is.na(beta_genes_cancer), arr.ind=TRUE)
beta_genes_cancer[l] <- rowMeans(beta_genes_cancer, na.rm=TRUE)[l[,1]]
```

#####Transforming beta-values into M-values

In the next step we **transform our beta-values into M-values** so we are able to do statistical tests with the data.

```{r}
M_genes_healthy <- log2(beta_genes_healthy/(1-(beta_genes_healthy)))
M_genes_cancer <- log2(beta_genes_cancer/(1-(beta_genes_cancer)))
```

#####Creating a dataset which contains M-values of sick and healthy patients

For apllying the PCA at our data we have to build a dataset which contains sick and healthy patients. 

```{r}
M_genes <- cbind(M_genes_healthy, M_genes_cancer)
```

###PCA

To **reduce our data** we use Principal Component Analysis. PCA reduces a large set of variables into a smaller set which still contains most information of the large dataset. Through the PCA correlated variables get transformed into a much smaller number of uncorrelated variables which are called promcipal components.
We want to have less variebles because it is easier to work with them. 

#####Code for the PCA

The Code for the **PCA** expects the patients to be rows and the samples to be columns. In our data the patients are columns and the genes are rows so we have to **transpose the matrix** with the t() function. 

```{r}
pca_M <- prcomp(t(M_genes))
```

#####How much variation do each component account for

To decide with how much principal components we want to work we have to know how much variation each principal component accounts for in percentage. We plot the percentage for each component and decide with the "elbow method" with wich number of copmonents we have to work with. 

```{r}
var_pca <- pca_M$sdev^2
var_pca_per <- round(var_pca/sum(var_pca)*100, 1)
plot(var_pca_per, main="variation of our data", xlab="Principal Components", ylab="Percent Variation", ylim=c(0,25),type = "o", pch=20)
```

#####Drawing a 2D plot of component 1 and 2

```{r}
library(ggplot2)

pca_values <- data.frame(Sample=rownames(pca_M$x),
                       X=pca_M$x[,1],
                       Y=pca_M$x[,2])
View(pca_values)

ggplot(data=pca_values, aes(x=X, y=Y, label=Sample)) +
  geom_text() +
  xlab(paste("PC1 - ", var_pca_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", var_pca_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PCA Graph")
```

#####Looking at the loading scores for the first principal component

Now we want to **find the 30 most important genes** which means we want to find the genes which have the most influence on the principal component one. For that we use the loading scores which we can find in rotation of the PCA. 

```{r}
loading_scores <- pca_M$rotation[,1]
gene_scores <- abs(loading_scores) 
ranked_gene_score <- sort(gene_scores, decreasing=TRUE)
genes_top_30 <- names(ranked_gene_score[1:30])
View(genes_top_30)
```

After that we want to see which genes have positive and which genes have negative loading scores.

```{r}
pca_M$rotation[genes_top_30,1]
```
